---
title: "Exploration European Invertebrate Freshwater Data"
author: "Raya Keuth"
date: 16.04.2024
output: 
  html_document:
    toc: yes
    toc-title: TOC
---
```{r, include=F}
setwd("/home/keuth/Documents/BMIP Workshop/BMIP Aquatic freshwater data/")

# Loading packages and data sets
library(RColorBrewer)
library(rworldmap)
library(scales)
library(maps)
library(stringr)
library(ggplot2)
library(tidyr)
library(dplyr)
library(tidyverse)
library(tibble)

# Load in data
Euro_Invert <- read.csv("data/Euro_FreshInv_all_sites.csv")
Euro_Invert_info_countries <- read.csv("data/Euro_FreshInv_information_country.csv")
Euro_Invert_info_studysites <- read.csv("data/Euro_FreshInv_information_studysite.csv")

# study countries
countries <- unique(Euro_Invert$country)
# study sites
study_sites <- unique(Euro_Invert$study_site)
```

The here-investigated data was published in the following paper [Haase et al. 2023](https://www.nature.com/articles/s41586-023-06400-1#Sec24).

## Spatial coverage

In total data from these 22 countries is available:

```{r, echo=FALSE}
# Plotting study countries (code obtained from the paper)
world_map <- maps::map("world", plot=FALSE)

ddf = read.table(text="
country
'France' 
'Estonia'
'Cyprus'
'N. Cyprus'
'Spain'
'Sweden'
'Luxembourg'
'Bulgaria'
'Denmark'
'Germany'
'United Kingdom'
'Ireland'
'Norway'
'Czech Rep.'
'Italy'
'Hungary'
'Portugal'
'Austria'
'Netherlands'
'Switzerland'
'Finland'
'Latvia'
'Belgium'", header=TRUE)

par(mar=c(0,0,0,0))

newmap <- getMap(resolution = "low")

col <- rep("grey85", length(newmap@data$NAME))
col[match(ddf$country, newmap@data$NAME)] <- c("black") #or pal instead of c("grey80")

col2 <- rep("#00000000", length(newmap@data$NAME)) #or alpha("white",0) instead of #00000000
col2[match(ddf$country, newmap@data$NAME)] <- c("grey50")

col2[58:59] <- c("black")#change borders for N.Cyprus
wv<-seq(1, 1, length.out=253) 
wv[58:59] <- c(0)

plot(newmap,col=col,
     bg="white",border=col2,
     xlim = c(-10, 34),
     ylim = c(35, 70),
     asp = 1,lwd=wv
)
```

In total 1816 different study sites were sampled. The study sites are unevenly distributed between the countries:
 
```{r, echo=FALSE}
# Plot number of study sites per country
ggplot(Euro_Invert_info_countries, aes(x=country, y= no_studysites))+
  geom_bar(stat = "identity")+
  labs(x = "", y = "")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 70, vjust = 1, hjust = 1)) +
  theme(axis.text.x = element_text(hjust = 1))
```

# Data exploration on the country level
## Number of records

```{r, echo = FALSE}
#Plot number of records (data points) per country
ggplot(Euro_Invert_info_countries, aes(x=country, y= no_records))+
  geom_bar(stat = "identity")+
  labs(x = "", y = "")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1)) +
  theme(axis.text.x = element_text(hjust = 0.5))+
  ggtitle("Number of records per country")

```

## Time coverage

The data set ranges from 1968 to 2020 (52 years). But not all years were sampled in every country:

```{r, echo=FALSE}
# Create list with the different countries as elements
Euro_Invert_list <- split(Euro_Invert, Euro_Invert$country)

# Sampled years in every country
sampling_years_countries <- lapply(Euro_Invert_list, function(x){unique(x$year)})

# data frame for completeness of time series for every year
completeness_timeseries_countries <- data.frame(Year = c(1968:2020))

# add a column for every study country
for (i in 1:length(countries)){
  completeness_timeseries_countries$tmp <- NA
  names(completeness_timeseries_countries)[names(completeness_timeseries_countries) == "tmp"] <- countries[i]
}

# obtain completeness of time series (non sampled years are marked as NA, sampled years as "Yes")
for (k in 1:length(countries)){
  for (i in 1968:2020) {
    if(i %in% sampling_years_countries[[countries[k]]] == T){
      completeness_timeseries_countries[completeness_timeseries_countries$Year == i, countries[k]] <- "Yes"
    }
  }
}

# Plotting completeness of time series for every country
# code for the plot modified from the visdat package 
# actual function: vis_dat(df)

# reorder dataframe to long format (kind of appending columns)
completeness_timeseries_countries_long <- completeness_timeseries_countries %>%
    pivot_longer(
      cols = -Year,
      names_to = "country",
      values_to = "sampled",
      values_transform = list(sampled = as.character)
    ) %>%
    arrange(Year, country, sampled)

# Proportion of sampled vs non-sampled
proportion_completeness_timeseries_countries_long <- data.frame(country = rep(Euro_Invert_info_countries$country,2), 
                                proportion = c(Euro_Invert_info_countries$p_years_miss, Euro_Invert_info_countries$p_years_pres),
                                type = c(rep("p_years_miss", nrow(Euro_Invert_info_countries)), rep("p_years_pres", nrow(Euro_Invert_info_countries))))

#Plot the completeness of sampled years per country
ggplot(data = completeness_timeseries_countries_long, aes(x = country, y = Year)) +
  geom_raster(aes(fill = sampled)) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "", y = "Year") +
  theme(axis.text.x = element_text(hjust = 0.5))+
  scale_fill_manual("Sampled?", values = "#FF6666", labels = c("Yes", "No"))+
  ggtitle("Coverage of sampled years per country")


# plot proportion as bar plot
ggplot(data = proportion_completeness_timeseries_countries_long, aes(x = country, y=proportion, fill = type))+
  geom_bar(position = "stack", stat = "identity")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "", y = "Proportion") +
  theme(axis.text.x = element_text(hjust = 0.5))+
  ggtitle("Proportion of missing/ sampled years per country")+
  scale_fill_manual("Proportion of", values = c("#C7C7C7", "#FF6A6A"), labels = c("Missing", "Sampled")) 
```

## Taxa coverage

In total 1402 species were found and 676 genus.

```{r, echo =F}
# Plot number of species per country
ggplot(Euro_Invert_info_countries, aes(x=country, y= no_species))+
  geom_bar(stat = "identity")+
  labs(x = "", y = "")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1)) +
  theme(axis.text.x = element_text(hjust = 0.5))+
  ggtitle("Number of identified species per country")

# Plot number of genus per country
ggplot(Euro_Invert_info_countries, aes(x=country, y= no_genus))+
  geom_bar(stat = "identity")+
  labs(x = "", y = "")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1)) +
  theme(axis.text.x = element_text(hjust = 0.5))+
  ggtitle("Number of identified genus per country")

```

Proportion of individuals identified to a specific taxonomic level. "Coarser" means that the individual was identified to sub-family-level or coarser.
```{r, echo =F}
# Proportion of taxon level

# Proportion of identified taxon level in every country
proportion_completeness_taxon_countries_long <- data.frame(country = rep(Euro_Invert_info_countries$country,3), 
                                proportion = c(Euro_Invert_info_countries$p_spec, Euro_Invert_info_countries$p_gen, Euro_Invert_info_countries$p_cors),
                                type = c(rep("per.species", nrow(Euro_Invert_info_countries)), rep("per.genus", nrow(Euro_Invert_info_countries)), 
                                         rep("per.coarser", nrow(Euro_Invert_info_countries))))

# plot results as bar plot
ggplot(data = proportion_completeness_taxon_countries_long, aes(x = country, y=proportion, fill = type))+
  geom_bar(position = "stack", stat = "identity")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1)) +
  labs(x = "", y = "Proportion") +
  theme(axis.text.x = element_text(hjust = 0.5))+
  ggtitle("Proportion of identified taxon level per country")+
  scale_fill_manual("Proportion of", values = c("#7FBADC", "#FFBF47", "#18848C"), labels = c("Coarser", "Species", "Genus"))
```

```{r,fig.width = 8, fig.height = 12, echo = FALSE, include = FALSE}
# Data exploration on the study site level
## Number of records


# Plot number of records per study site
ggplot(Euro_Invert_info_studysites, aes(x=study_sites, y= no_records))+
  geom_bar(stat = "identity")+
  labs(x = "", y = "")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1)) +
  theme(axis.text.x = element_text(hjust = 1), strip.text = element_text(size = 12), plot.title = element_text(size=18))+
  ggtitle("Number of records per study site")+
  facet_wrap(~country, scales = "free_x")

```



```{r,fig.width = 10, fig.height = 12, echo=FALSE, include= FALSE}
## Time coverage

# Create list with the different study sites as elements
Euro_Invert_list_studysite <- split(Euro_Invert, Euro_Invert$study_site)

# Sampled years in every study site
sampling_years_studysite <- lapply(Euro_Invert_list_studysite, function(x){unique(x$year)})

# data frame for completeness of time series for every year
completeness_timeseries_studysite <- data.frame(Year = c(1968:2020))

# add a column for every study site
for (i in 1:length(study_sites)){
  completeness_timeseries_studysite$tmp <- NA
  names(completeness_timeseries_studysite)[names(completeness_timeseries_studysite) == "tmp"] <- study_sites[i]
}

# obtain completeness of time series (non sampled years are marked as NA, sampled years as "Yes")
for (k in 1:length(study_sites)){
  for (i in 1968:2020) {
    if(i %in% sampling_years_studysite[[study_sites[k]]] == T){
      completeness_timeseries_studysite[completeness_timeseries_studysite$Year == i, study_sites[k]] <- "Yes"
    }
  }
}

# reorder dataframe to long format (kind of appending columns)
completeness_timeseries_studysites_long <- completeness_timeseries_studysite %>%
  pivot_longer(
    cols = -Year,
    names_to = "study_sites",
    values_to = "sampled",
    values_transform = list(sampled = as.character)
  ) %>%
  arrange(Year, study_sites, sampled)

completeness_timeseries_studysites_long <- as.data.frame(completeness_timeseries_studysites_long)

# add column with country
completeness_timeseries_studysites_long$country <- NA
for (k in 1:length(countries)) {
    row_index <- grep(countries[k],completeness_timeseries_studysites_long[,"study_sites"])
    completeness_timeseries_studysites_long[row_index, "country"] <- countries[k]
}

# Proportion of sampled/ non-sampled countries
proportion_completeness_timeseries_studysites_long <- data.frame(study_site = rep(Euro_Invert_info_studysites$study_site,2), 
                                proportion = c(Euro_Invert_info_studysites$p_years_miss, Euro_Invert_info_studysites$p_years_pres),
                                type = c(rep("p.miss", nrow(Euro_Invert_info_studysites)), rep("p.pres", nrow(Euro_Invert_info_studysites))))

# add column with country
proportion_completeness_timeseries_studysites_long$country <- NA
for (k in 1:length(countries)) {
  row_index <- grep(countries[k], proportion_completeness_timeseries_studysites_long[,"study_site"])
  proportion_completeness_timeseries_studysites_long[row_index, "country"] <- countries[k]
}

# Plot results of completeness
ggplot(data = completeness_timeseries_studysites_long, aes(x = study_sites, y = Year)) +
  geom_raster(aes(fill = sampled)) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1)) +
  labs(x = "", y = "Year") +
  theme(axis.text.x = element_text(hjust = 1), strip.text = element_text(size = 12), plot.title = element_text(size=18))+
  scale_fill_manual("Sampled?", values = "#FF6666", labels = c("Yes", "No"))+
  ggtitle("Coverage of sampled years per study site")+
  facet_wrap(~country, scales = "free_x")

# plot proportion as bar plot
ggplot(data = proportion_completeness_timeseries_studysites_long, aes(x = study_site, y=proportion, fill = type))+
  geom_bar(position = "stack", stat = "identity")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1)) +
  labs(x = "", y = "Proportion") +
  theme(axis.text.x = element_text(hjust = 1), strip.text = element_text(size = 12), plot.title = element_text(size=18))+
  ggtitle("Proportion of missing/ sampled years per study site")+
  scale_fill_manual("Proportion of", values = c("#C7C7C7", "#FF6A6A"), labels = c("Missing", "Sampled"))+
  facet_wrap(~country, scales = "free_x")
```


```{r,fig.width = 8, fig.height = 12, echo =F, include= FALSE}
## Taxa coverage

# Plot number of species per study site
ggplot(Euro_Invert_info_studysites, aes(x=study_sites, y= no_species))+
  geom_bar(stat = "identity")+
  labs(x = "", y = "")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1)) +
  theme(axis.text.x = element_text(hjust = 1), strip.text = element_text(size = 12), plot.title = element_text(size=18))+
  ggtitle("Number of species per study site")+
  facet_wrap(~country, scales = "free_x")

# Plot number of genus per study site
ggplot(Euro_Invert_info_studysites, aes(x=study_sites, y= no_genus))+
  geom_bar(stat = "identity")+
  labs(x = "", y = "")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1)) +
  theme(axis.text.x = element_text(hjust = 1), strip.text = element_text(size = 12), plot.title = element_text(size=18))+
  ggtitle("Number of genus per study site")+
  facet_wrap(~country, scales = "free_x")

```


```{r,fig.width = 10, fig.height = 12, echo =F, include = FALSE}
#Proportion of individuals identified to a specific taxonomic level. "Coarser" means that the individual was identified to sub-family-level or coarser.
# Proportion of taxon level
# change data frame to long format
proportion_completeness_taxon_studysites_long <- data.frame(study_site = rep(Euro_Invert_info_studysites$study_site,3), 
                                      proportion = c(Euro_Invert_info_studysites$p_spec, Euro_Invert_info_studysites$p_gen, Euro_Invert_info_studysites$p_cors),
                                      type = c(rep("per.species", nrow(Euro_Invert_info_studysites)), rep("per.genus", nrow(Euro_Invert_info_studysites)), 
                                               rep("per.coarser", nrow(Euro_Invert_info_studysites))))

# add column with country
proportion_completeness_taxon_studysites_long$country <- NA
for (k in 1:length(countries)) {
  row_index <- grep(countries[k], proportion_completeness_taxon_studysites_long[,"study_site"])
  proportion_completeness_taxon_studysites_long[row_index, "country"] <- countries[k]
}

# plot results as bar plot
ggplot(data = proportion_completeness_taxon_studysites_long, aes(x = study_site, y=proportion, fill = type))+
  geom_bar(position = "stack", stat = "identity")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1)) +
  labs(x = "", y = "Proportion") +
  theme(axis.text.x = element_text(hjust = 1), strip.text = element_text(size = 12), plot.title = element_text(size=18))+
  ggtitle("Proportion of identified taxon level per study site")+
   scale_fill_manual("Proportion of", values = c("#7FBADC", "#FFBF47", "#18848C"), labels = c("Coarser", "Species", "Genus"))+
  facet_wrap(~country, scales = "free_x")
```
